%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SB07 -- Structural Evolution}

\subsection*{Table of Contents}

\begin{enumerate}
\item Protein evolution
\item Sequence-structure relationship
\item Structural conservation
\item Homology and analogy
\item Divergent evolution
\item Convergent evolution
\item Protein folds and fold space
\item Structural classification systems
\item Remote homology detection
\item Evolutionary constraints on protein structure
\end{enumerate}

\subsection*{Possible Exam Questions:}

\begin{enumerate}
\item{Why is protein structure generally more conserved than sequence? \done}
\item{\textcolor{blue}{What is the relation between sequence similarity and structural similarity in biological proteins?}\done}
\item{What are homologous proteins? Explain the difference between paralogues and orthologues \done}
\item{What is a protein fold?}
\item{Explain the Rossmann fold as an example of structural conservation.}
\item{How does evolution constrain protein structure?}
\item{Why is structural information useful for detecting remote homology?}
\end{enumerate}


\paragraph{Why is protein structure generally more conserved than sequence?}
Because the three-dimensional structure of a protein is what determines its function, there is strong evolutionary pressure to maintain the structure. The sequence can tolerate more mutations as long as the overall fold and function of the protein are preserved. 

\paragraph{What is the relation between sequence similarity and structural similarity in biological proteins?}
Protein structure is generally more conserved than sequence in evolution.
Proteins that share $>\sim30\%$ of sequence identity almost always have significantly similar
folds. Below this threshold ($20-30\%$ - the "twilight zone"), structural similarity can still exist despite low sequence similarity.
Summarising: similar sequence implies similar structure, but similar structure does not imply
similar sequence.

\paragraph{What are homologous proteins? Explain the difference between paralogues and orthologues}
Homologous proteins are proteins that are believed to derive from a common evolutionary ancestor \textit{gene}. Homologous proteins can be classified into two categories: orthologues and paralogues.
\begin{itemize}
\item \textbf{Orthologues} are homologous proteins found in different species that originated by a \textbf{speciation event}. Since they usually retain the same biological role, orthologues often have similar functions. For example, the same enzyme in human and mouse is typically an orthologue.
\item \textbf{Paralogues} are homologous proteins that originated by a gene duplication event within a genome (within a single species). After duplication, the two copies start to accumulate mutations and diverge, and the two proteins may acquire different functions. Therefore, paralogues are evolutionarily related, but they are not necessarily functionally equivalent. An example is the genes that do olfactory recognition: they differentiated from a commn ancestor gene and then each specialized in recognition of different olfactory molecules.
\end{itemize}
The term "homologous" must not be confused with "analogous", which instead defines proteins that have similar structure (and, sometimes, also similar function) but do not derive from a common ancestor. 


\paragraph{How does evolution constrain protein structure?}

Evolution constrains protein structure because a protein sequence is accepted only if it can fold into a \textit{stable}, functional structure in a \textit{reasonable} time.
Evolution also favours some folds over others. This can be understood as \textbf{fold competition}: different folds can in principle act as scaffolds for similar functions, but some folds are better because they position active-site or binding residues more effectively while remaining stable and easy to fold. These folds are therefore selected and reused more often. In addition, the number of possible protein architectures is limited by the available \textbf{supersecondary structures}, such as $\beta$-hairpins, helix-turn-helix motifs, and $\beta\alpha\beta$ motifs. Since only some combinations of helices, strands, and loops are physically stable and fold efficiently, evolution repeatedly reuses a limited set of successful structural solutions.